He-Man: A domain-specific language for lightweight concurrency

### 2011-12-13 report/report.tex
Cleaned up miscellaneous things in report.

\begin{abstract}
We present He-Man, a Haskell compiled, embedded, domain specific
language for building scalable network applications in a natural,
straight-line manner while still emitting efficient event-loop based
code. He-Man programs are compact and measurements indicate that it
performs acceptably. 
\end{abstract}

@@

An improvement is to write multithreaded code, in which each thread serves
multiple clients. This is the most popular approach, and is utilized in the
Apache HTTP server \cite{Apache}.

@@

@@

The underlying abstract syntax for the front-end He-Man language is a
fairly conventional imperative, statement based, first order language,
shown in Figure \ref{fig:front-syntax}. The notable constructs are
\b{Spawn}, which creates a new thread that takes a list of arguments
and launches the associated code with those arguments available to it,
and \b{Wait}, which takes an event expression as an argument and
blocks until the event has ``completed''. Note that \b{Call} is for
calling foreign functions; He-Man does not have its own notion of
function.

Since writing programs using the abstract syntax would be highly
inconvenient, we provide a monadic library for writing He-Man
programs, which allows us to leverage Haskell's do notation to write
He-Man programs using convenient imperative syntax. In this way, we
are taking advantage of monads being ``programmable semi-colons''.

@@

At the expression level, \t{Expr} is an instance of the \t{Num} type
class so that numeric literals and operations can be used in He-Man
syntax. We also provide infix operators for other useful operations on
expressions, some of them with mildly awkward names, since the good
names were already taken by a function with the wrong type.

The result of all of this is that He-Man code can be written in a
fairly natural style, as shown in Figure \ref{fig:code-ex}.

@@

As an example of the power that Haskell and the monadic style gives
us, Figure \ref{fig:generic} shows our library routines for simulating
blocking network operations. \t{do\_nb\_action} takes an event
\t{Expr} and a \t{Prog Expr} action representing a network call that
can return \t{EAGAIN} and produces an action that generates code to
simulate a blocking network operation. This allows us to describe how
to simulate blocking I/O in a generic way, and then invoke it to
create our concrete wrappers.
(\t{.=.} is a wrapper for
\b{Assign} that takes a \t{Prog Expr} as its second argument and the
\t{sock\_*} functions invoke \b{Call} wrappers on the appropriate
network calls).

@@

accept (fd, e) =
  do_nb_action e (sock_accept fd)
do_read (fd, e) buf size =
  do_nb_action e (sock_read fd buf size)
do_write (fd, e) buf size =
  do_nb_action e (sock_write fd buf size)
\end{smallverbatim}
\caption{Implementation of ``blocking'' network calls}
\label{fig:generic}
\end{figure}

@@

This pass generates a correct state machine, albeit one which contains many
blocks with only tail-position jumps. We follow it up with the \t{optimizeJumps}
pass, which removes these empty blocks, redirecting all \t{Goto}s to their
ultimate destinations, and allowing nested \t{If}s in tail position. Nested
\t{If}s occur frequently in generated code, for example, immediately before an
\t{if} statement at the end of a \t{while} loop, when the code might next jump
inside the \t{if} statement, to the top of the \t{while} loop, or after the
\t{while} loop, depending on multiple tests. Although this pass is, strictly
speaking, an optimization, it is critical for obtaining high-quality output.

At this point, the program's control flow is correct, but many blocks will
contain uses of variables which have been declared in previous blocks. The
\t{collectFrees} pass traverses each block, collecting the variables used but
not declared in each block. (Recall that our front-end ensures that a variable
is only ever declared in one location, so there is no ambiguity.) The local
variables of each thread are the union of the used-not-declared variables in
each of its blocks. We save this list of local variables in the appropriate
\i{Thread} container.

Note that we could simply mark all declared variables as thread-local, but
variables which are only used in same block as their declaration need not be
stored thread-locally. We want to minimize the number of thread-local variables,
because this directly impacts the memory overhead of each thread.

@@

\t{Stmt}s are translated straightforwardly; threads are spawned by allocating a
new \t{struct} for the corresponding thread, setting its continuation to the
first block in the thread, and adding it to the list of runnable threads.
\t{Expr}s and \t{Tail}s are similarly straightforward; variables in the thread
\t{struct} must be replaced by references to their thread-local value, while
other variables are guaranteed to be block-local. Finally, we must also generate
a \t{main} function which spawns the first thread and jumps to the main loop.

@@

The He-Man runtime main loop is very simple, then: it loops
indefinitely, and on each iteration it removes the thread at the head
of the runqueue (if it exists) and runs it, and polls for IO events.

The runtime provides a blocking mechanism based on an event
abstraction. To the runtime, an event consists of a pointer to a
thread that is blocked on it and some arbitrary identifier (in
practice the file descriptor). To create an event for use with
nonblocking IO, an event is allocated, marked with the associated file
descriptor, and a new edge-triggered \t{epoll} event is created for the
file descriptor using \t{epoll\_ctl}, with the event structure as the
associated pointer data.
We use edge-triggered instead of level-triggered so that we can wait
on both read and write events for a given file descriptor without
having to adjust which we are waiting for at runtime, and because
He-Man's programming model makes it easy to meet the requirement that
blocking on a file descriptor is only done after an operation on the
descriptor has returned \t{EAGAIN}.
The actual polling for IO events is done by calling \t{epoll}. For each
event that has been returned by epoll, if a thread is waiting on the
event, the thread is added to the runqueue. If the the runqueue is
empty, the call to epoll is blocking, otherwise it is called with a
timeout of zero seconds, to simply poll for events.

@@

To compare the ease of writing event based network applications, we
built simple echo servers using both He-Man and C. The C version was
written in as an event loop driven explicit state machine using the
He-Man runtime. The C version, which is fairly tightly written, is 98
lines of source, according to SLOCCount. It is fairly readable, but
its structure is obscured by being split into states and the need to
explicitly handle read and write returning EAGAIN and write returning
short. The state machine has a total of five states. The He-Man
version is 33 lines long and is a very clear representation of the
control flow of each individual thread. (Additionally, the He-Man
library has 17 lines of code implementing generic routines for
simulating blocking IO using nonblocking IO and wait.) The total line
counts, however, somewhat understate the difference between the
versions, since the boilerplate for setting up connections is about
the same for both versions. When just looking at the core read/write
loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

@@

Our benchmark measurements are taken using ab, the Apache HTTP server
benchmarking tool. \cite{ApacheAB} Measurements were collected by
having a variable number of client threads request a file 100 times;
the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
file. We ran the tests against both our He-Man http server and Apache
2.2.20, installed with the default Ubuntu 11.10 settings, except for
client, thread, and process limits being adjusted upwards. Since the
He-Man runtime is currently only single threaded, we ran the Apache
web server pinned to a single processor. Since the test repeatedly
requests the same file, it will always be in the kernel's disk cache,
and thus this test is not disk bound.

For our experiments, the server ran Linux kernel 3.0.0 on a Lenovo
Thinkpad W520 with a quad core 2.3 GHz Intel Core i7-2820QM and 8
gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
Intel Core 2 Duo and 8 gigabytes of RAM.  % XXX: client specs
The machines were connected by a gigabit Ethernet link.

Unfortunately, ab is not particularly scalable, so we had trouble
testing with more clients than 500.

@@

\begin{figure*}[!htb]
	\centering
	\includegraphics[width=0.9\textwidth]{128kb_graph.png}
	\caption{Web server benchmark, 128kb file}
	\label{fig:graph128}
\end{figure*}
\begin{figure*}[!htb]
	\centering
	\includegraphics[width=0.9\textwidth]{16kb_graph.png}
	\caption{Web server benchmark, 16kb file}
	\label{fig:graph16}
\end{figure*}

@@

The per-thread memory overhead imposed by He-Man is very small: it
consists of a thread id, list pointers for the scheduling queue, the
next state to enter, a pointer to the most recently received event,
and list heads and tails for lists of the events being waited on,
events owned by the thread, and buffers owned by the thread (the
latter two are so that we can automatically reclaim these resources
when the thread exits). On a 64-bit system, the total overhead comes
out to 88 bytes, and could easily be reduced further (e.g. by using
singly linked lists).  For comparison, the default thread stack size
under NPTL on Linux is 2 megabytes.

>>>>

\begin{abstract}
We present He-Man, a compiled, embedded, domain specific language in Haskell for
building scalable network applications in a natural, straight-line manner while
still producing efficient event loop-based code. He-Man programs are compact and
measurements indicate that it performs acceptably. Our implementation is
available at
\url{https://github.com/msullivan/networks-project}.
\end{abstract}

@@

An improvement is to write multithreaded code in which each thread serves
multiple clients. This is the most popular approach, and is utilized in the
Apache HTTP server \cite{Apache}.

@@

The latest version of He-Man can be found at
\url{https://github.com/msullivan/networks-project}.

@@

The underlying abstract syntax for the front-end He-Man language is a
fairly conventional imperative, statement-based, first-order language,
shown in Figure \ref{fig:front-syntax}. The notable constructs are
\b{Spawn}, creating a new thread, which takes a list of arguments
and launches the associated code with those arguments available to it,
and \b{Wait}, which takes an event expression as an argument and
blocks until the event has ``completed''. Note that \b{Call} is for
calling foreign functions; He-Man does not have its own notion of
function.

Since writing programs using the abstract syntax would be highly
inconvenient, we provide a monadic library for writing He-Man
programs, which allows us to leverage Haskell's do notation to write
He-Man programs using convenient imperative syntax. In this way, we
are taking advantage of monads as ``programmable semicolons''.

@@

At the expression level, \t{Expr} is an instance of the \t{Num} typeclass 
so that numeric literals and operations can be used in He-Man
syntax. We also provide infix operators for other useful operations on
expressions, some of them with mildly awkward names to avoid name conflicts.

As a result, He-Man code can be written in a fairly natural style with little
need for normal Haskell code, as shown in Figure \ref{fig:code-ex}.

@@

As an example of the power that Haskell and the monadic style gives
us, Figure \ref{fig:generic} shows our library routines for simulating
blocking network operations. \t{do\_nb\_action} takes an event
\t{Expr} and a \t{Prog Expr} action representing a network call that
can return \t{EAGAIN} and produces an action that generates code to
simulate a blocking network operation. This allows us to describe how
to simulate blocking I/O in a generic way, and then invoke it to
create our concrete wrappers. (\t{.=.} is a wrapper for \b{Assign} 
that takes a \t{Prog Expr} as its second argument and the \t{sock\_*} 
functions invoke \b{Call} wrappers on the appropriate network calls.)

@@

accept (fd, e) =
  do_nb_action e (sock_accept fd)
do_read (fd, e) buf size =
  do_nb_action e (sock_read fd buf size)
do_write (fd, e) buf size =
  do_nb_action e (sock_write fd buf size)
\end{smallverbatim}
\caption{Implementation of ``blocking'' network calls.}
\label{fig:generic}
\end{figure}

@@

This pass generates a correct state machine, albeit one which contains many
blocks with only tail-position jumps. We follow it up with the \t{optimizeJumps}
pass, which removes these empty blocks, redirecting all \b{Goto}s to their
ultimate destinations, and allowing nested \b{If}s in tail position. Nested
\b{If}s occur frequently in generated code, for example, immediately before an
\t{if} statement at the end of a \t{while} loop, when the code might next jump
inside the \t{if} statement, to the top of the \t{while} loop, or after the
\t{while} loop, depending on multiple tests. Although this pass is, strictly
speaking, an optimization, it is critical for obtaining high-quality output.

At this point, the program's control flow is correct, but many blocks will
contain uses of variables which have been declared in previous blocks. The
\t{collectFrees} pass traverses each block, collecting the variables used but
not declared in each block. (Our front-end ensures that a variable is only ever
declared in one location, so there is no ambiguity.) The local variables of each
thread are the union of the used-not-declared variables in each of its blocks.
We save this list of local variables in the appropriate \i{Thread} container.

Note that we could simply mark all declared variables as thread-local, but
variables which are only used in same block as their declaration need not be
retained outside that block. We want to minimize the number of thread-local
variables, because this directly impacts the memory overhead of each thread.

@@

\i{Stmt}s are translated straightforwardly; threads are spawned by allocating a
new \t{struct} for the corresponding thread, setting its continuation to the
first block in the thread, and adding it to the list of runnable threads.
\i{Expr}s and \i{Tail}s are similarly straightforward; variables in the thread
\t{struct} must be replaced by references to their thread-local value, while
other variables are guaranteed to be block-local. Finally, we must also generate
a \t{main} function which spawns the first thread and jumps to the main loop.

@@

The He-Man runtime main loop is very simple, then: it loops
indefinitely, and on each iteration it removes the thread at the head
of the runqueue (if it exists) and runs it, and polls for I/O events.

The runtime provides a blocking mechanism based on an event
abstraction. To the runtime, an event consists of a pointer to a
thread that is blocked on it and some arbitrary identifier (in
practice, the file descriptor). To create an event for use with
nonblocking I/O, an event is allocated, marked with the associated file
descriptor, and a new edge-triggered \t{epoll} event is created for the
file descriptor using \t{epoll\_ctl}, with the event structure as the
associated pointer data.
We use edge-triggered instead of level-triggered so that we can wait
on both read and write events for a given file descriptor without
having to adjust which we are waiting for at runtime, and because
He-Man's programming model makes it easy to meet the requirement that
blocking on a file descriptor is only done after an operation on the
descriptor has returned \t{EAGAIN}.
The actual polling for I/O events is done by calling \t{epoll}. For each
event that has been returned by \t{epoll}, if a thread is waiting on the
event, the thread is added to the runqueue. If the the runqueue is
empty, the call to \t{epoll} is blocking, otherwise it is called with a
timeout of zero seconds, to simply poll for events.

@@

To compare the ease of writing event based network applications, we
built simple echo servers using both He-Man and C. The C version was
written in as an event loop driven explicit state machine using the
He-Man runtime. The C version, which is fairly tightly written, is 98
lines of source, according to SLOCCount. It is fairly readable, but
its structure is obscured by being split into states and the need to
explicitly handle read and write returning \t{EAGAIN} and write returning
short. The state machine has a total of five states. The He-Man
version is 33 lines long and is a very clear representation of the
control flow of each individual thread. (Additionally, the He-Man
library has 17 lines of code implementing generic routines for
simulating blocking IO using nonblocking IO and wait.) The total line
counts, however, somewhat understate the difference between the
versions, since the boilerplate for setting up connections is about
the same for both versions. When just looking at the core read/write
loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

@@

Our benchmark measurements are taken using ab, the Apache HTTP server
benchmarking tool \cite{ApacheAB}. Measurements were collected by
having a variable number of client threads request a file 100 times;
the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
file. We ran the tests against both our He-Man HTTP server and Apache
2.2.20, installed with the default Ubuntu 11.10 settings, except for
client, thread, and process limits being adjusted upwards. Since the
He-Man runtime is currently only single-threaded, we ran the Apache
web server pinned to a single processor. Since the test repeatedly
requests the same file, it will always be in the kernel's disk cache,
and thus this test is not disk bound.

For our experiments, the server ran Linux kernel 3.0.0 on a Lenovo
Thinkpad W520 with a quad core 2.3 GHz Intel Core i7-2820QM and 8
gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
Intel Core 2 Duo and 8 gigabytes of RAM.  % XXX: client specs
The machines were connected by a gigabit Ethernet link.
Unfortunately, ab is not particularly scalable, so we had trouble
testing with more than 500 clients.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{128kb_graph.png}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{16kb_graph.png}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}

@@

The per-thread memory overhead imposed by He-Man is very small: it
consists of a thread id, list pointers for the scheduling queue, the
next state to enter, a pointer to the most recently received event,
and list heads and tails for lists of the events being waited on,
events owned by the thread, and buffers owned by the thread (the
latter two are so that we can automatically reclaim these resources
when the thread exits). On a 64-bit system, the total overhead comes
out to 88 bytes, and could easily be reduced further (e.g., by using
singly-linked lists).  For comparison, the default thread stack size
under NPTL on Linux is 2 megabytes.
